---
title: "Darwin Core mapping"
subtitle: "For: RINSE - Pathways and vectors of biological invasions in Northwest Europe"
author:
- Lien Reyserhove
- Peter Desmet
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how we map the checklist data to Darwin Core. The source file for this document can be found [here](https://github.com/trias-project/rinse-pathways-checklist/blob/master/src/dwc_mapping.Rmd).

# Setup

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Set locale (so we use UTF-8 character encoding):

```{r}
# This works on Mac OS X, might not work on other OS
Sys.setlocale("LC_CTYPE", "en_US.UTF-8")
```

Load libraries:

```{r}
library(tidyverse) # To transform data
library(magrittr)  # For %<>% pipes
library(janitor)   # To clean input data
library(readxl)    # To read Excel files
library(stringr)   # To perform string operations
library(digest)    # To generate hashes
library(rgbif)     # To parse names
library(purrr)     # To clean references
```

Set file paths (all paths should be relative to this script):
 
```{r}
# Raw files:
raw_data_file = "../data/raw/transcribed_10530_2016_1278_MOESM2_ESM.xlsx"
references = "../data/raw/transcribed_references.xlsx"

# Processed files:
dwc_taxon_file = "../data/processed/taxon.csv"
dwc_literature_references_file = "../data/processed/references.csv"
dwc_distribution_file = "../data/processed/distribution.csv"
dwc_profile_file = "../data/processed/speciesprofile.csv"
dwc_description_file = "../data/processed/description.csv"
```

# Read and pre-process raw data

Create a data frame `raw_data` from the source data:

```{r}
raw_data <- read_excel(path = raw_data_file) 
```

Clean the data somewhat: remove empty rows if present and clean names:

```{r}
raw_data %<>% 
  remove_empty("rows") %<>%
  clean_names()
```

## Generate taxonID

To uniquely identify a taxon in the taxon core and reference taxa in the extensions, we need a `taxonID`. Since we need it in all generated files, we generate it here in the raw data frame. It is a combination of `dataset-shortname:taxon:` and a hash based on the scientific name. As long as the scientific name doesn't change, the ID will be stable. Some of the names require some minor cleaning first though:

```{r}
raw_data %<>% mutate(species_clean = case_when(
  species == "Crassostrea rhizophoraeGuilding 1828" ~ "Crassostrea rhizophorae Guilding 1828",
  species == "Petricola pholadiformisLamarck, 1818" ~ "Petricola pholadiformis Lamarck, 1818",
  species == "Theba pisanaMüller, 1774)" ~ "Theba pisana (Müller, 1774)",
  species == "Maylandia sp. M. K. Meyer & W. Förster, 1984" ~ "Maylandia M. K. Meyer & W. Förster, 1984",
  TRUE ~ species)) 
```

Some scientific names don't have spaces before the `(`. We add them here:

```{r}
raw_data %<>% mutate(species_clean = str_replace_all(species_clean, "\\(", " ("), 
                     species_clean = str_replace_all(species_clean, "  ", " "))
```

Generate `taxonID`:

```{r}
# Vectorize the digest function (the digest() function isn't vectorized. So if you pass in a vector, you get one value for the whole vector rather than a digest for each element of the vector):
vdigest <- Vectorize(digest)

# Generate taxonID:
raw_data %<>% mutate(taxonID = paste("rinse-pathways-checklist", "taxon", vdigest (species_clean, algo = "md5"), sep = ":"))
```

## Further pre-processing

Add prefix `raw_` to all column names to avoid name clashes with Darwin Core terms:

```{r}
colnames(raw_data) <- paste0("raw_", colnames(raw_data))
```

Preview data:

```{r}
raw_data %>% head()
```

# Create taxon core

```{r start_taxon}
taxon <- raw_data
```

## Term mapping
 
Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).
 
### language

```{r}
taxon %<>% mutate(language = "en")
```

### license 

```{r}
taxon %<>% mutate(license = "http://creativecommons.org/publicdomain/zero/1.0/") 
```

### rightsHolder

```{r}
taxon %<>% mutate(rightsHolder = "University of Cambridge")
```

### accessRights

```{r}
taxon %<>% mutate(accessRights = "http://www.inbo.be/en/norms-for-data-use") 
```

### datasetID

```{r}
taxon %<>% mutate(datasetID = "")
```

### institutionCode

```{r}
taxon %<>% mutate(institutionCode = "University of Cambridge") 
```

### datasetName

```{r}
taxon %<>% mutate(datasetName = "RINSE - Pathways and vectors of biological invasions in Northwest Europe")
```

### taxonID

```{r}
taxon %<>% mutate(taxonID = raw_taxonID)
```

### scientificName

```{r}
taxon %<>% mutate(scientificName = raw_species_clean)
```

### kingdom

Information for the higher classification is contained in `raw_higher_classification`:

```{r}
taxon %>% distinct(raw_higher_classification)
```

These map to different taxonomical levels (e.g. `Mammalia` = class while `Osteichthyes` = order). We interpret this information in the mapping of `kingdom`, `phylum`, `class` and `order`. 

```{r}
taxon %<>% mutate(kingdom = case_when(
  raw_higher_classification == "Angiospermae" ~ "Plantae" ,
  TRUE ~ "Animalia")) 
```

### phylum

```{r}
taxon %<>% mutate(phylum = case_when(
  raw_higher_classification == "Angiospermae" ~ "Magnoliophyta" ,
  raw_higher_classification == "Mollusca" ~ "Mollusca",
  TRUE ~ "Chordata")) 
```

### class

Osteichthyes, which is a superclass, includes Actinopterygii and Sarcopterygii (https://en.wikipedia.org/wiki/Osteichthyes). All the taxa in this list are Actinopterygii which is a class.

```{r}
taxon %<>% mutate(class = case_when(
  raw_higher_classification == "Anseriformes" ~ "Aves",
  raw_higher_classification == "Mammalia" ~ "Mammalia",
  raw_higher_classification == "Osteichthyes" ~ "Actinopterygii", 
  TRUE ~ "")) 
```

### order

```{r}
taxon %<>% mutate(order = case_when(
  raw_higher_classification == "Anseriformes" ~ "Anseriformes",
  TRUE ~ "" )) 
```

### taxonRank

The taxon rank can be provided by the `nameparser()` function of rgbif. First we create a data frame with the parsed names:

```{r gbif_parse_names}
taxon_ranks <- parsenames(taxon$raw_species_clean)
```

And then select rank information from `rankmarker`. The field `notho` contains valuable information for the generic hybrids, so this column is selected as well. 

```{r}
taxon_ranks %<>% select(scientificname, rankmarker, notho) 
```

Merge `taxon_ranks` with `taxon` using scientific names:

```{r}
taxon %<>% left_join(taxon_ranks, by = c("raw_species_clean" = "scientificname")) 
```

Show unique values for `rankmarker` and `notho`:

```{r}
taxon %>% 
  select(rankmarker, notho) %>% 
  group_by_all() %>% 
  summarise(records = n())
```

Show `scientificName`s for which `rankmarker` = `NA`:

```{r}
taxon %>% 
  select(scientificName, rankmarker) %>% 
  filter(is.na(rankmarker))
```

`Maylandia M. K. Meyer & W. Förster, 1984` is a genus, the other two records are species.

Map `taxonRank`:

```{r}
taxon %<>% mutate(taxonRank = case_when(
  !is.na(notho) ~ "hybrid",
  rankmarker == "sp." & is.na(notho) ~ "species",
  rankmarker == "infrasp." ~ "subspecies",
  is.na(rankmarker) & scientificName != "Maylandia M. K. Meyer & W. Förster, 1984" ~ "species",
  scientificName == "Maylandia M. K. Meyer & W. Förster, 1984" ~ "genus")) 
```

## Post-processing

Remove the original columns:

```{r}
taxon %<>% select(-starts_with("raw_"), -rankmarker, - notho)
```

Sort on `taxonID`:

```{r}
taxon %<>% arrange(taxonID)
```

Preview data:

```{r}
taxon %>% head()
```

Save to CSV:

```{r}
write.csv(taxon, file = dwc_taxon_file, na = "", row.names = FALSE, fileEncoding = "UTF-8")
```

# Create literature reference extension

```{r}
literature_references <- raw_data
```

## Pre-processing

Information for the literature references is contained in `raw_references`, which are (sequences of) numbers in this case. The link between these numbers and the full references is provided in the original [Supplementary material file](https://github.com/trias-project/rinse-pathways-checklist/blob/master/data/raw/10530_2016_1278_MOESM2_ESM.docx) (page 28-30). We copy-pasted this information to a newly generated Excel file [references](https://github.com/trias-project/rinse-pathways-checklist/blob/master/data/raw/transcribed_references.xlsx), here saved as the dataframe `transcribed_references`:

```{r}
transcribed_references <- read_excel(path = references, col_types = "text")
```

### Full references (on per row)

The literature reference extension is a dataframe in which one row contains one single reference (a short identifier and the full reference) per taxon. We generate this file in several steps:

1. Provide **all** reference numbers in `raw_references`, instead of using the format containing `-` for a range of numbers between a delimited minimun and maximum (e.g. `3-7` is transformed to `3,4,5,6,7`). To accomplish this, we first create the function `hyphen_to_sequence()`:

```{r}
hyphen_to_sequence <- function(input){
  seq_locs <- gregexpr("\\d+-\\d+", input)
  matched <- regmatches(input, seq_locs)[[1]]
  
  split_support <- function(text) {
    split_seq <- strsplit(text, "-")[[1]]
    paste(as.character(seq(split_seq[1], split_seq[2])), collapse = ",")
  }
  
  regmatches(input, seq_locs, invert = FALSE) <- list(map_chr(matched, split_support))
  input
}
```

Apply function:

```{r}
literature_references %<>% 
  rowwise() %<>%
  mutate(raw_references_clean = hyphen_to_sequence(raw_references)) 
```

2. Transform `literature_references` from a wide to a long dataset, with each row containing one reference number per taxon:

```{r}
literature_references %<>%
  mutate(references_sep = raw_references_clean) %<>% 
  ungroup() %<>%
  separate_rows(references_sep, sep = ",")
```

3. Replace the reference numbers provided in `raw_references` with the full references provided in the dataframe `transcribed_references` using the field `number` as a link. 

```{r}
literature_references %<>% left_join(transcribed_references, by = c("references_sep" = "number"))
```

This dataframe constitutes the basis for further mapping of the literature reference extension:

```{r}
literature_references %>%
  select(raw_species, identifier, full_reference) %>%
  head()
```

### Source references (sequence of identifiers)

For the mapping of the field `source` in the other extensions, we need to translate the sequence of reference numbers given for each taxon in `raw_references` into a sequence of all corresponding identifiers separated by ` | `. As we need to do this several times, we generate a new dataframe `identifier_sequence` here, linking the sequence of numbers in `raw_references` with the sequence of identifiers.

Generate dataframe `identifier_sequence` with the required columns (add taxonID as this is necessary for spreading in the next step): 

```{r}
identifier_sequence <- literature_references %>% select(raw_taxonID, raw_references, references_sep, identifier)
```

To generate the required sequence of identifiers, we need to transform `identifier_sequence` from a dataframe containing one identifier per taxon per row, to one containing a sequence of identifiers per taxon per row. We do this by using the `spread()` function first, generating 33 identifier columns (corresponding to the 33 identifiers):

```{r}
identifier_sequence %<>% mutate(references_sep = as.numeric(references_sep)) %<>% 
  spread(references_sep, identifier) %<>%
  select(-`<NA>`)
```

In the following step, we unite these 33 columns into one field `id_sequence`, using ` | ` as a separator. To unite the identifier columns (`1-33`) in a single step, we first rename the column names by adding the prefix `ref_`:

```{r}
old_names <- c("raw_taxonID", "raw_references", c(1:33))
new_names <- c("raw_taxonID", "raw_references", paste0("ref_", c(1:33)))
identifier_sequence %<>% rename_at(vars(old_names), ~ new_names)
```

Generate `id_sequence`:

```{r}
identifier_sequence %<>% unite("id_sequence", c(paste0("ref_", c(1:33))), sep = " | ")
```

Preview:

```{r}
identifier_sequence %>% head()
```

Remove `NA`'s from the generated sequences:

```{r}
identifier_sequence %<>% mutate(id_sequence = str_replace_all(id_sequence, "(([NA]{2}\\s[|]) | [|]\\s([NA]{2}))", ""))
```

Remove `raw_taxonID`:

```{r}
identifier_sequence %<>% select(-raw_taxonID) 
```

Remove duplicate rows (some taxa have the same reference sequence):

```{r}
identifier_sequence %<>% distinct(raw_references, .keep_all = TRUE)
```

We will use this dataframe to link the sequence of references numbers (`raw_references`) with the sequence of identifiers (`id_sequence`) in the mapping of the following extensions.

## Term mapping

Map the data to [Literature References](http://rs.gbif.org/extension/gbif/1.0/references.xml).

### taxonID

```{r}
literature_references %<>% mutate(taxonID = raw_taxonID)
```

### identifier

```{r}
literature_references %<>% select(-identifier, everything()) 
```

### bibliographicCitation

```{r}
literature_references %<>% mutate(bibliographicCitation = full_reference) 
```

## Post-processing

Remove the original columns:

```{r}
literature_references %<>% select(-starts_with("raw_"), -references_sep, -full_reference)
```

Sort on `taxonID`:

```{r}
literature_references %<>% arrange(taxonID) 
```

Preview data:

```{r}
literature_references %>% head()
```

Save to CSV:

```{r}
write.csv(literature_references, file = dwc_literature_references_file, na = "", row.names = FALSE, fileEncoding = "UTF-8")
```

# Create distribution extension

```{r start_distribution}
distribution <- raw_data
```

## Pre-processing

Information for the distributions and associated dates of first observation is contained in eight columns in total:

- 4 columns for the distribution information (Great Britain, France, Belgium and the Netherlands in resp. `raw_established_gb`, `raw_established_france`, `raw_established_belgium` and `raw_established_netherlands`)
- 4 columns for the first record of the taxon in these countries (resp. `raw_first_record_gb`, `raw_first_record_france`, `raw_first_record_belgium` and `raw_first_record_netherlands`). 

When a taxon is present in a country, this is indicated by a `Y` and a date of first record is provided.

We need to reduce the information for eight to two columns: one column containing country presence information (`country column`, which is the basis for `locality`, `locationID` and `countryCode`) and column containing the date of first observance (`date` column, which is the basis for `eventDate`). We do this in three steps:

1. Paste date and establishment information for each country into one column (from 8 to 4 columns):

```{r}
distribution %<>% mutate(record_gb = paste(raw_established_gb, raw_first_record_gb, sep = ","))  %<>% 
  mutate(record_fr = paste(raw_established_france, raw_first_record_france, sep = ",")) %<>% 
  mutate(record_be = paste(raw_established_belgium, raw_first_record_belgium, sep = ",")) %<>%
  mutate(record_nl = paste(raw_established_netherlands, raw_first_record_netherlands, sep = ",")) 
```

2. Generate `country` and `date` by gathering the new columns (from 4 to 2 columns):

```{r}
distribution %<>% gather("country", "date", record_gb, record_fr, record_be, record_nl) 
```

3. Clean `date` by removing records with `NA,NA` and removing `Y`:

```{r}
distribution %<>% filter(date != "NA,NA") %<>% mutate(date = str_replace_all(date, "Y,", "")) 
```

4. Rename `country`:

```{r}
distribution %<>% mutate(country = recode(country,
  "record_gb" = "Great Britain",
  "record_fr" = "France",
  "record_be" = "Belgium",
  "record_nl" = "The Netherlands")
)
```

Preview:

```{r}
distribution %>%
  select(raw_species, country, date) %>%
  head()
```

## Term mapping

Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml).

### taxonID

```{r}
distribution %<>% mutate(taxonID = raw_taxonID)
```

### locationID

```{r}
distribution %<>% mutate(locationID = case_when(
   country == "Great Britain" ~ " WGSRPD:GRB",
   country == "France" ~ "ISO_3166-2:FR",
   country == "Belgium" ~ "ISO_3166-2:BE",
   country == "The Netherlands" ~ "ISO_3166-2:NL"))
```

### locality

```{r}
distribution %<>% mutate(locality = country)
```

### countryCode

```{r}
distribution %<>% mutate(countryCode = case_when(
   country == "Great Britain" ~ "GB",
   country == "France" ~ "FR",
   country == "Belgium" ~ "BE",
   country == "The Netherlands" ~ "NL"))
```

### occurrenceStatus

```{r}
distribution %<>% mutate(occurrenceStatus = "present") 
```

### establishmentMeans

```{r}
distribution %<>% mutate(establishmentMeans = "introduced") 
```

### eventDate

Inspect dates:

```{r}
distribution %>% distinct(date) %>% arrange(date)
```

Check which dates deviate from the required ISO 8601 format (here `YYYY`)

```{r}
unique(distribution[which(str_detect(distribution $ date,"^[0-9]{4}$") == FALSE), "date"])
```

Clean those dates:

```{r}
distribution %<>% mutate(date = str_trim(date)) # remove whitespaces

distribution %<>% mutate(date = recode(date,    # recode deviating formats
   "n.d."  = "",
   "17634" = "1763",
   "2004," = "2004",
   "n..d"  = "",
   "199"   = "1990",
   "1985," = "1985",
   "NA"    = "",
   "19204" = "1904")) 
```

After this cleaning step, there should be no deviations from the ISO 8601 format:

```{r}
unique(distribution[which(str_detect(distribution $ date,"^[0-9]{4}$") == FALSE), "date"])
```

We want to express `eventDate` as the range between the date of first and last record. In this case, we do not have information about the last record. We will consider the publication year of Zieritz et al. (2017) as the date when the presence of the species was last verified:

```{r}
distribution %<>% mutate(eventDate = case_when(
  date == "" ~ "",
  date != "" ~ paste(date, "2017", sep = "/"))) 
```

### source

Information about the source is provided in `raw_references`. We replace the sequence of reference numbers by the sequence of identifiers (not full references), provided in the dataframe `identifier_sequence` generated earlier.

```{r}
distribution %<>% left_join(identifier_sequence, by = ("raw_references"))
```

Rename `id_sequence`:

```{r}
distribution %<>%  rename("source" = "id_sequence")
```

## Post-processing

Remove the original columns:

```{r}
distribution %<>% select(-starts_with("raw_"), -country, -date)
```

Sort on `taxonID`:

```{r}
distribution %<>% arrange(taxonID) 
```

Preview data:

```{r}
distribution %>% head()
```

Save to CSV:

```{r}
write.csv(distribution, file = dwc_distribution_file, na = "", row.names = FALSE, fileEncoding = "UTF-8")
```

# Create species profile extension

In this extension we express broad habitat characteristics (e.g. `isTerrestrial`) of the species.

```{r start_species_profile}
species_profile <- raw_data
```

Habitat information can be found in `raw_environment`, which describes whether a species is found in freshwater, terrestrial, marine or a combination of these habitats.

Show unique values:

```{r}
species_profile %>%
  select(raw_environment) %>%
  group_by_all() %>% 
  summarize(records = n())
```

Only two taxa have no information on the environment:

- `Salvelinus fontinalis` (brook trout): a fish found in both freshwater and marine habitats
- `Rattus norvegicus`(brown rat): a terrestrial mammal

We will add this information in the mapping. Environment information is mapped to `isFreshwater`, `isTerrestrial`, `isMarine` or a combination of these terms in the species profile extension.

## Term mapping

Map the data to [Species Profile](http://rs.gbif.org/extension/gbif/1.0/speciesprofile.xml).

### taxonID

```{r}
species_profile %<>% mutate(taxonID = raw_taxonID)
```

### isMarine

```{r}
species_profile %<>% mutate(isMarine = case_when(
  raw_environment == "M" |
  raw_environment == "M+F" |
  raw_environment == "T+M" | 
  raw_species == "Salvelinus fontinalis (Mitchill, 1814)" ~ "TRUE",
  TRUE ~ "FALSE"
))
```

### isFreshwater

```{r}
species_profile %<>% mutate(isFreshwater = case_when(
  raw_environment == "F" |
  raw_environment == "F,T" |
  raw_environment == "F+T" |
  raw_environment == "M+F" |
  raw_species == "Salvelinus fontinalis (Mitchill, 1814)" ~ "TRUE",
  TRUE ~ "FALSE"))
```

### isTerrestrial

```{r}
species_profile %<>% mutate(isTerrestrial = case_when(
  raw_environment == "F,T" |
  raw_environment == "F+T" |
  raw_environment == "T" |
  raw_environment == "T+M" |
  raw_species == "Rattus norvegicus (Berkenhout 1769)" ~ "TRUE",
  TRUE ~"FALSE"))
```

Show mapped values:

```{r}
species_profile %>%
  select(raw_environment, isMarine, isFreshwater, isTerrestrial) %>%
  group_by_all() %>%
  summarize(records = n())
```

## Post-processing

Remove the original columns:

```{r}
species_profile %<>% select(-starts_with("raw_"))
```

Sort on `taxonID`:

```{r}
species_profile %<>% arrange(taxonID)
```

Preview data:

```{r}
species_profile %>% head()
```

Save to CSV:

```{r}
write.csv(species_profile, file = dwc_profile_file, na = "", row.names = FALSE, fileEncoding = "UTF-8")
```

# Create description extension

In the description extension we want to include several important characteristics (hereafter refered to as descriptors) about the species:

- Native range
- Pathway of introduction

A single taxon can have multiple descriptions of the same type (e.g. multiple native ranges), expressed as multiple rows in the description extension. For each descriptor, we create a separate dataframe to process the specific information. We always specify _which descriptor_ we map (`type` column) and its _specific content_ (`description` column). After the mapping of these Darwin Core terms `type` and `value`, we merge the dataframes to generate one single description extension. We then continue the mapping process by adding the other Darwin Core terms (which content is independent of the type of descriptor, such as `language`).

## Native range

Native range information (e.g. `Australia`) can be found in `raw_origin`. 

Create separate dataframe:

```{r}
native_range <- raw_data
```

Show unique values:

```{r}
native_range %>% 
  distinct(raw_origin) %>% 
  arrange(raw_origin)
```

`raw_origin` contains multiple values (currently not more than 4), so we separate it in 4 columns:

```{r}
native_range %<>% separate(raw_origin, into = paste("origin", c(1:4), sep = "_"))
```

Gather in a `key` and `value` column:
 
```{r}
native_range %<>% gather("key", "value", paste("origin", c(1:4), sep = "_"), na.rm = TRUE) 
```

Inspect values:

```{r}
native_range %>% 
  distinct(value) %>% 
  arrange(value)
```

Map values:

```{r}
native_range %<>% mutate(mapped_value = recode(value,
  "Af"  = "Africa (WGSRPD:2)",
  "Ar"  = "Arctic",
  "As"  = "Asia",
  "Au"  = "Australia (WGSRPD:50)",
  "Aus" = "Australia (WGSRPD:50)",
  "d"   = "",
  "Eu"  = "Europe (WGSRPD:1)",
  "n"   = "",
  "na"  = "",
  "NAm" = "Northern America (WGSRPD:7)",
  "SAm" = "Southern America (WGSRPD:8)"
)) 
```

Show mapped values:

```{r}
native_range %>%
  select(value, mapped_value) %>%
  group_by(value, mapped_value) %>%
  summarize(records = n()) 
```

Keep only non-empty descriptions:

```{r}
native_range %<>% filter(!is.na(mapped_value) & mapped_value != "")
```

Drop the `key` and `value` columns and rename `mapped_value` as `description`:

```{r}
native_range %<>%
  select(-key, -value) %>%
  rename(description = mapped_value)
```

Create a `type` field to indicate the type of description:

```{r}
native_range %<>% mutate(type = "native range")
```

## Pathway of introduction

Pathway information in this dataset is a combination of the columns starting with `raw_pathway_` (4 columns: `raw_pathway_accidental`, `raw_pathway_disperal`, `raw_pathway_import_escape` and `raw_pathway_import_release`) and `raw_vector_` (5 columns: `raw_vector_biocontrol`, `raw_vector_leisure`, `raw_vector_industry`, `raw_vector_ornamental` and `raw_vector_research`):

```{r}
raw_data %>% select(starts_with("raw_vector"), starts_with("raw_pathway")) %>% 
  group_by_all() %>% 
  summarize(records = n()) 
```

This information will be mapped to the six **main categories** of the [CBD standard](https://www.cbd.int/doc/meetings/sbstta/sbstta-18/official/sbstta-18-09-add1-en.pdf). How these RINSE pathways map to the CBD standard is described in the source paper Zieritz et al. (2017) (in Methods > Data analysis):

RINSE pathway | CBD category 
--- | ---
accidental | stowaway, contaminant
dispersal | corridor, undaided
import_release | release
import_escape | escape

For certain taxa, we will complement this pathway information with the specific vector of introduction (contained in columns starting with `raw_vector_`), using the **subcategories** of the CBD standard:

- For pathways `accidental` and `disperal`, we will **never** provide vector information. This is because vectors of accidentally introduced species were not analysed in the present study, due to a lack of reliable data (see Zieritz et al. 2017).
- For pathways `import_escape` and `import_release`, we will map vector information to the CBD subcategories **only** when they map unambigiously to this standard (see table below and [this issue](https://github.com/trias-project/rinse-pathways-checklist/issues/3)) **and** when these vectors are the only vectors of introduction for these taxon.

RINSE pathway | RINSE vector | CBD category | CBD subcategory
--- | --- | --- | ---
accidental | biocontrol | stowaway, contaminant | NA
accidental | industry | stowaway, contaminant | NA
accidental | leisure | stowaway, contaminant | NA
accidental | ornamental | stowaway, contaminant | NA
accidental | research | stowaway, contaminant | NA
dispersal | biocontrol | corridor, unaided | NA
dispersal | industry | corridor, unaided | NA
dispersal | leisure | corridor, unaided | NA
dispersal | ornamental | corridor, unaided | NA
dispersal | research | corridor, unaided | NA
import_escape | biocontrol | escape | NA
import_escape | industry | escape | NA
import_escape | leisure | escape | live food and bait
import_escape | ornamental | escape | NA 
import_escape | research | escape | research and ex-situ breeding (in facilities)
import_release | biocontrol | release  | biological control
import_release | industry | release | NA
import_release | leisure | release | NA
import_release | ornamental | release | NA
import_release | research | release | NA

We map this information in the new column `description`, for which we use the [TrIAS vocabulary](https://github.com/trias-project/vocab/blob/master/vocabulary/pathway/pathways.tsv).

Create separate dataframe:

```{r}
pathway_desc <- raw_data
```

We gather pathway information in one new column `pathway`:

```{r}
pathway_desc %<>% gather(pathway, value, starts_with("raw_pathway_"), na.rm = TRUE, convert = TRUE)
```

Column "value" will only contain "Y", so no need for this column

```{r}
pathway_desc %<>% select(-value)
```

Map patwhay information to the CBD main category:

```{r}
pathway_desc %<>% mutate(description = recode(pathway,
  "raw_pathway_accidental" = "stowaway,contaminant",
  "raw_pathway_dispersal" = "corridor,natural_dispersal",
  "raw_pathway_import_escape" = "escape",
  "raw_pathway_import_release" = "release"
))
```

Arrange by taxonID to see things more in context:

```{r}
pathway_desc %<>% arrange(raw_taxonID)
```

```{r echo=FALSE}
pathway_desc %>% select(raw_taxonID, description) %>% head(n = 10)
```

Map vector information to CBD subcategories for some taxa:

```{r}
pathway_desc %<>% mutate(description = case_when(
  description == "escape" &
    raw_vector_research == "Y" & 
    raw_vector_leisure == "Y" &
    is.na(raw_vector_ornamental) &
    is.na(raw_vector_biocontrol) & 
    is.na(raw_vector_industry) ~ "escape_research, escape_food_bait",
  description == "escape" &
    raw_vector_research == "Y" & 
    is.na(raw_vector_leisure ) &
    is.na(raw_vector_ornamental) &
    is.na(raw_vector_biocontrol) & 
    is.na(raw_vector_industry) ~ "escape_research",
  description == "escape" &
    is.na(raw_vector_research) & 
    raw_vector_leisure == "Y" &
    is.na(raw_vector_ornamental) &
    is.na(raw_vector_biocontrol) & 
    is.na(raw_vector_industry) ~ "escape_food_bait",
  description == "release" &
    is.na(raw_vector_research) & 
    is.na(raw_vector_leisure) &
    is.na(raw_vector_ornamental) &
    raw_vector_biocontrol == "Y" & 
    is.na(raw_vector_industry) ~ "release_biocontrol",
TRUE ~ description # Leave rest as is
))
```

Separate `description` when more than two values are provided, e.g. in case of RINSE pathway = `accidental` or RINSE pathway = `dispersal`:

```{r}
pathway_desc %<>% separate_rows(description, sep = ",")
```

Remove all empty description values:

```{r}
pathway_desc %<>% filter(description != "") 
```

Remove duplicate rows:

```{r}
pathway_desc %<>% distinct(raw_taxonID, description, .keep_all = TRUE)
```

Remove `pathway`:

```{r}
pathway_desc %<>% select(-pathway)
```

Add the prefix `cbd_2014_pathway:` to refer to this standard:

```{r}
pathway_desc %<>% mutate(description = paste ("cbd_2014_pathway", description, sep = ":"))
```

Create a `type` field to indicate the type of description:

```{r}
pathway_desc %<>% mutate(type = "pathway")
```

## Union descriptions

Union native range and pathway of introduction:

```{r start_description_ext}
description_ext <- bind_rows(native_range, pathway_desc)
```

## Term mapping
 
Map the data to [Taxon Description](http://rs.gbif.org/extension/gbif/1.0/description.xml).

### taxonID

```{r}
description_ext %<>% mutate(taxonID = raw_taxonID)
```

### description

```{r}
description_ext %<>% mutate(description = description)
```

### type

```{r}
description_ext %<>% mutate(type = type)
```

### source

Information about the source is provided in `raw_references`. We replace the sequence of reference numbers by the sequence of identifiers (not full references), provided in the dataframe `identifier_sequence` generated earlier.

```{r}
description_ext %<>% left_join(identifier_sequence, by = ("raw_references"))
```

Rename `id_sequence`:

```{r}
description_ext %<>%  rename("source" = "id_sequence")
```

### language

```{r}
description_ext %<>% mutate(language = "en")
```

## Post-processing

Remove the original columns:

```{r}
description_ext %<>% select(-starts_with("raw_"))
```

Move `taxonID` to the first position:

```{r}
description_ext %<>% select(taxonID, everything())
```

Sort on `taxonID`:

```{r}
description_ext %<>% arrange(taxonID)
```

Preview data:

```{r}
description_ext %>% 
  mutate(source = substr(source, 1, 20)) %>% # Shorten source to make it easier to display
  head()
```

Save to CSV:

```{r}
write.csv(description_ext, file = dwc_description_file, na = "", row.names = FALSE, fileEncoding = "UTF-8")
```
