---
title: "Darwin Core mapping"
subtitle: "For: RINSE - pathways and vectors of biological invasions in Northwest Europe"
author:
- Lien Reyserhove
- Peter Desmet
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how we map the checklist data to Darwin Core. The source file for this document can be found [here](https://github.com/trias-project/alien-plants-belgium/tree/master/data/raw).

# Setup

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Set locale (so we use UTF-8 character encoding):

```{r}
# This works on Mac OS X, might not work on other OS
Sys.setlocale("LC_CTYPE", "en_US.UTF-8")
```

Load libraries:

```{r}
library(tidyverse) # To transform data
library(magrittr)  # For %<>% pipes
library(janitor)   # To clean input data
library(readxl)    # To read Excel files
library(stringr)   # To perform string operations
library(digest)    # To generate hashes
library(rgbif)     # Interface to the GBIF API
```

Set file paths (all paths should be relative to this script):
 
```{r}
# Raw files:
raw_data_file = "../data/raw/copy_of_10530_2016_1278_MOESM2_ESM.xlsx"

# Processed files:
dwc_taxon_file = "../data/processed/taxon.csv"
dwc_distribution_file = "../data/processed/distribution.csv"
dwc_profile_file = "../data/processed/speciesprofile.csv"
dwc_description_file = "../data/processed/description.csv"
```

# Read and pre-process raw data

Create a data frame `raw_data` from the source data:

```{r}
raw_data <- read_excel(path = raw_data_file) 
```

Clean the data somewhat: remove empty rows if present and clean names:

```{r}
raw_data %<>% remove_empty_rows() %<>%     # Remove empty rows
  clean_names()
```

